Dominance Regions for Imaginary $\bfg$-vectors

### 2021-09-07 dominance_regions.tex
Some notation changes and other edits

\newcommand{\ol}[1]{{\overline{#1}}}

@@

  Observe that the data of $D$ is just the data of a bipartition of a finite totally ordered set $D=D_1\sqcup D_2$.
  Given a Dyck path $D=D_1\sqcup D_2$ in the lattice rectangle $[0,a_1]\times[0,a_2]$, write $\overline{D}=\overline{D}_1\sqcup \overline{D}_2$ for the Dyck path in the lattice rectangle $[0,a_2]\times[0,a_1]$ with $\overline{D}_1=a_1+a_2+1-D_2$ and $\overline{D}_2=a_1+a_2+1-D_1$.
  This operation can be visualized as transposing the Dyck path.

  Given any quantity or object $P$ whose definition depends on $b$ and $c$, write $P^\vee$ for the same quantity or object defined with $b$ and $c$ interchanged.
  In particular, we have $u^\vee_{m,\varepsilon}=u_{m,-\varepsilon}$ for any $m$ and any $\varepsilon=\pm1$.

@@

  As an immediate consequence, we determine the structure of certain maximal Dyck paths.
  We introduce the following notation for convenience:
  \[d_m:=\begin{cases} b & \text{if $m$ is even;}\\ c & \text{if $m$ is odd;}\end{cases} \qquad \text{and} \qquad d^\vee_m:=\begin{cases} c & \text{if $m$ is even;}\\ b & \text{if $m$ is odd;}\end{cases}\]
  The following was proven in \cite[Corollary 2.6]{rupel} and follows by a simple induction using Lemma~\ref{le:maximal Dyck path recursion}.
  \begin{corollary}
    Assume $b,c\ge2$ (this can be dropped with extra notation and slight modification).
    Then, for $\varepsilon=\pm1$ and $m\ge1$, the maximal Dyck path $D_{m,\varepsilon}:=D^{max}[u_{m,\varepsilon},u_{m-1,-\varepsilon}]$ can be constructed as follows:
    \begin{itemize}
      \item $D_{1,\varepsilon}$ consists of a single horizontal edge;
      \item $D_{2,\varepsilon}$ consists of $d_2$ horizontal edges followed by a vertical edge;
      \item for $m\ge3$, $D_{m,\varepsilon}$ consists of $d_m-1$ copies of $D_{m-1,\varepsilon}$ followed by a copy of $D_{m-1,\varepsilon}$ with its first $D_{m-2,\varepsilon}$ removed;
    \end{itemize}
    and the maximal Dyck path $\overline{D}^\vee_{m,\varepsilon}:=D^{max}[u_{m-1,\varepsilon},u_{m,-\varepsilon}]$ can be constructed as follows:
    \begin{itemize}
      \item $\overline{D}^\vee_{1,\varepsilon}$ consists of a single vertical edge;
      \item $\overline{D}^\vee_{2,\varepsilon}$ consists of a horizontal edge followed by $d^\vee_2$ vertical edges;
      \item for $m\ge3$, $\overline{D}^\vee_{m,\varepsilon}$ consists of a copy of $\overline{D}^\vee_{m-1,\varepsilon}$ with its last $\overline{D}^\vee_{m-2,\varepsilon}$ removed followed by $d^\vee_m-1$ copies of $\overline{D}^\vee_{m-1,\varepsilon}$.
    \end{itemize}
  \end{corollary}

  For $e \le e'\in D$, write $ee'$ for the subpath of $D$ beginning with $e$ and ending with $e'$.
  When $e=e'$, we take $ee'$ to be the subpath consisting of only the edge $e$.
  \begin{definition}
    For $H\sqcup V\subset D$ with $H\subset D_0$ and $V\subset D_2$ together with a subpath $ee'\subset D$, define the \emph{horizontal shadow statistic}:
    \begin{equation}
      \label{eq:horizontal shadow statistic}
      f_H(ee')=-|ee'\cap D_2|+c|ee'\cap H|
    \end{equation}
    and the \emph{vertical shadow statistic}:
    \begin{equation}
      \label{eq:vertical shadow statistic}
      f_V(ee')=-|ee'\cap D_1|+b|ee'\cap V|.
    \end{equation}
  \end{definition}
  \begin{remark}
    Observe that the shadow statistics are additive with respect to concatenation of subpaths.
  \end{remark}
  \begin{definition}
    A subset $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$ is \emph{compatible} if: for every $h\in H$ and $v\in V$ with $h<v$, there exists an edge $e\in hv$ such that one of the following conditions is satisfied
    \[\tag{HGC} e\ne v \quad \text{and} \quad f_H(he)=0\]
    \[\tag{VGC} e\ne h \quad \text{and} \quad f_V(ev)=0\]
    Write $\cC_D$ for the collection of all compatible subsets of $D$.
  \end{definition}
  \begin{remark}
    In the construction of the greedy basis, it is necessary to consider $D$ as closed cycle by identifying $(0,0)$ with $(a_1,a_2)$ and allowing subpaths $ee'$ with $e'<e$ to ``wrap around'' $D$ in the obvious way.
    When we wish to emphasize that this condition should be imposed, we denote the Dyck path $D$ instead by $D_\WA$.
    For $(a_1,a_2)=(u_{m+1,+},u_{m,-})$ or $(u_{m,+},u_{m+1,-})$, the wrap around condition does not restrict compatibility and can thus safely be omitted.
  \end{remark}

@@

  \begin{lemma}
    Consider $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$.
    \begin{enumerate}
      \item For $h\in H$, take $he$ to be either the path $D(h;H)$ or $D_\WA(h;H)$.
        The following hold:
        \begin{enumerate}
          \item for any proper subpath $he'\subsetneq he$, we have $f_H(he')>0$;
          \item for any proper subpath $e'e\subsetneq he$, we have $f_H(e'e)<0$;
          \item $e\in V$.
        \end{enumerate}
      \item For $v\in V$, take $ev$ to be either the path $D(v;V)$ or $D_\WA(v;V)$.
        The following hold:
        \begin{enumerate}
          \item for any proper subpath $e'v\subsetneq ev$, we have $f_V(e'v)>0$;
          \item for any proper subpath $ee'\subsetneq ev$, we have $f_V(ee')<0$;
          \item $e\in H$.
        \end{enumerate}
    \end{enumerate}
  \end{lemma}
  \begin{proof}
    We prove (1), the proof of (2) is similar.
    To see (a), note that as $e'$ traverses the path $he$ from $h$ to $e$, the value of $f_H(he')$ either increases by $c$, stays constant, or decreases by 1.
    Since $h\in H$, we have $f_H(hh)=c>0$ and so the value of $f_H(he')$ stays positive until it first becomes 0 when $e'=e$.
    Part (b) then follows from the additivity of $f_H$ and part (c) is the case when $e'=e$ here.
  \end{proof}

@@

    For $a_1,a_2\in\ZZ_{\ge0}$ with $(-a_2,-a_1+ca_2)\in\cI'$, define the \emph{dual dominant Dyck path} $D^{dom^\vee}=D^{dom^\vee}[a_1,a_2]$ to be the maximal Dyck path starting from $(0,0)$ and terminating at $(a_1,a_2)$ while staying below the lines $y=\frac{bc+\sqrt{bc(bc-4)}}{2c}x$ and $y-a_2=\frac{bc-\sqrt{bc(bc-4)}}{2c}(x-a_1)$.
  \end{definition}
  \begin{remark}
    \label{rem:intersection}
    The lines defining $D^{dom}[a_1,a_2]$ meet at the point 
    \[m_{a_1,a_2}:=\left( \frac{2ba_2-\big(bc-\sqrt{bc(bc-4)}\big)a_1}{2\sqrt{bc(bc-4)}} , \frac{\big(bc+\sqrt{bc(bc-4)}\big)a_2-2ca_1}{2\sqrt{bc(bc-4)}} \right).\]
  \end{remark}

@@

  \begin{theorem}
    \label{th:dominant Dyck path recursion}
    For $a_1,a_2\in\ZZ_{\ge0}$ with $(-a_1+ba_2,-a_2)\in\cI$, we have $\mu_{V,b}\big(D^{dom}[a_1,a_2]\big)=D^{dom^\vee}[a_2,ba_2-a_1]$ and $\mu_{H,c}\big(D^{dom}[a_1,a_2]\big)=D^{dom^\vee}[ca_1-a_2,a_1]$.
  \end{theorem}
  \begin{proof}
    We prove the result for $\mu_{H,c}$ as the result for $\mu_{V,b}$ immediately follows from $\mu_{H,b}$ being its inverse.
    Write $D'$ for the Dyck path $\mu_{H,c}\big(D^{dom}[a_1,a_2]\big)$ and observe that by construction $D'$ has $ca_1-a_2$ horizontal edges and $a_1$ vertical edges.
    Write $v'_1,\ldots,v'_{a_1}$ for the vertical edges of $D'$ with $v'_r$ immediately preceded by $\ell_r$ horizontal edges.
    We have to show (i) that every lattice point on $D'$ stays below the lines 
    \[y=\frac{bc+\sqrt{bc(bc-4)}}{2c}x \qquad \text{and} \qquad y-a_1=\frac{bc-\sqrt{bc(bc-4)}}{2c}\big(x-(ca_1-a_2)\big)\]
    while (ii) any lattice point above $D'$ is also above one of these lines.

>>>>

\newcommand{\ol}[1]{{\overline{#1}}}
\newcommand{\vv}[1]{{{}^\vee \! #1}}

@@

  Observe that the data of $D$ is just the data of a bipartition of a finite totally ordered set $D=D_1\sqcup D_2$.
  Given a Dyck path $D=D_1\sqcup D_2$ in the lattice rectangle $[0,a_1]\times[0,a_2]$, write $\bar{D}=\bar{D}_1 \sqcup \bar{D}_2$ for the Dyck path in the lattice rectangle $[0,a_2]\times[0,a_1]$ with $\bar{D}_1=a_1+a_2+1-D_2$ and $\bar{D}_2=a_1+a_2+1-D_1$.
  This operation can be visualized as transposing the Dyck path.

  Given any quantity or object $P$ whose definition depends on $b$ and $c$, write $\vv{P}$ for the same quantity or object defined with $b$ and $c$ interchanged.
  In particular, we have $\vv{u}_{m,\varepsilon}=u_{m,-\varepsilon}$ for any $m$ and any $\varepsilon=\pm1$.

@@

  As an immediate consequence, we determine the structure of certain maximal Dyck paths.
  We introduce the following notation for convenience:
  \[d_m:=\begin{cases} b & \text{if $m$ is even;}\\ c & \text{if $m$ is odd;}\end{cases} \qquad \text{and} \qquad \vv{d}_m:=\begin{cases} c & \text{if $m$ is even;}\\ b & \text{if $m$ is odd;}\end{cases}\]
  The following was proven in \cite[Corollary 2.6]{rupel} and follows by a simple induction using Lemma~\ref{le:maximal Dyck path recursion}.
  \begin{corollary}
    Assume $b,c\ge2$ (this can be dropped with extra notation and slight modification).
    Then, for $\varepsilon=\pm1$ and $m\ge1$, the maximal Dyck path $D^{m,\varepsilon}:=D^{max}[u_{m,\varepsilon},u_{m-1,-\varepsilon}]$ can be constructed as follows:
    \begin{itemize}
      \item $D^{1,\varepsilon}$ consists of a single horizontal edge;
      \item $D^{2,\varepsilon}$ consists of $d_2$ horizontal edges followed by a vertical edge;
      \item for $m\ge3$, $D_{m,\varepsilon}$ consists of $d_m - 1$ copies of $D^{m-1,\varepsilon}$ followed by a copy of $D^{m-1,\varepsilon}$ with its first $D^{m-2,\varepsilon}$ removed;
    \end{itemize}
    and the maximal Dyck path $\vv{\bar{D}}^{m,\varepsilon}:=D^{max}[u_{m-1,\varepsilon},u_{m,-\varepsilon}]$ can be constructed as follows:
    \begin{itemize}
      \item $\vv{\bar{D}}^{1,\varepsilon}$ consists of a single vertical edge;
      \item $\vv{\bar{D}}^{2,\varepsilon}$ consists of a horizontal edge followed by $\vv{d}_2$ vertical edges;
      \item for $m\ge3$, $\vv{\bar{D}}^{m,\varepsilon}$ consists of a copy of $\vv{\bar{D}}^{m-1,\varepsilon}$ with its last $\vv{\bar{D}}^{m-2,\varepsilon}$ removed followed by $\vv{d}_m - 1$ copies of $\vv{\bar{D}}^{m-1,\varepsilon}$.
    \end{itemize}
  \end{corollary}

  \subsection{Compatibility}
  For $e \le e'\in D$, write $ee'$ for the subpath of $D$ beginning with $e$ and ending with $e'$.
  When $e=e'$, we take $ee'$ to be the subpath consisting of only the edge $e$.
  \begin{definition}
    For $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$ together with a subpath $ee'\subset D$, define the \emph{horizontal shadow statistic}:
    \begin{equation}
      \label{eq:horizontal shadow statistic}
      f_H(ee')=-|ee'\cap D_2|+c|ee'\cap H|
    \end{equation}
    and the \emph{vertical shadow statistic}:
    \begin{equation}
      \label{eq:vertical shadow statistic}
      f_V(ee')=-|ee'\cap D_1|+b|ee'\cap V|.
    \end{equation}
  \end{definition}
  \begin{remark}
    Observe that the shadow statistics are additive with respect to concatenation of subpaths.
  \end{remark}
  \begin{definition}
    Consider $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$.
    For $h\in H$ and $v\in V$ with $h<v$, call the path $hv$ \emph{compatible} if there exists an edge $e\in hv$ such that one of the following conditions is satisfied:
    \[\tag{HGC} e\ne v \quad \text{and} \quad f_H(he)=0;\]
    \[\tag{VGC} e\ne h \quad \text{and} \quad f_V(ev)=0.\]
    Call the subset $H\sqcup V$ \emph{compatible} if every path $hv$ with $h\in H$ and $v\in V$ is compatible.
    Write $\cC_D$ for the collection of all compatible subsets of $D$.
    Given a subset $H\subset D_1$, write $\cC_D(H)$ for the collection of all $V\subset D_2$ for which $H\sqcup V\in\cC_D$.
    Define $\cC_D(V)$ for $V\subset D_2$ similarly.
  \end{definition}
  \begin{remark}
    In the construction of the greedy basis, it is necessary to consider $D$ as a closed cycle by identifying $(0,0)$ with $(a_1,a_2)$ and allowing subpaths $ee'$ with $e'<e$ to ``wrap around'' $D$ in the obvious way.
    When we wish to emphasize that this condition should be imposed, we denote the Dyck path $D$ instead by $D_\WA$.
    Following \cite[Remark 2.21]{rupel}, for $D^{m,\varepsilon}$ and $\vv{\bar{D}}^{m,\varepsilon}$ the wrap around condition does not restrict compatibility and can thus safely be omitted.
  \end{remark}

@@

  \begin{lemma}
    \label{le:shadow statistics}
    Consider $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$.
    \begin{enumerate}
      \item For $h\in H$, take $he$ to be either the path $D(h;H)$ or $D_\WA(h;H)$.
        The following hold:
        \begin{enumerate}
          \item for any proper subpath $he'\subsetneq he$, we have $f_H(he')>0$;
          \item for any proper subpath $e'e\subsetneq he$, we have $f_H(e'e)<0$;
          \item $e\in V$.
        \end{enumerate}
      \item For $v\in V$, take $ev$ to be either the path $D(v;V)$ or $D_\WA(v;V)$.
        The following hold:
        \begin{enumerate}
          \item for any proper subpath $e'v\subsetneq ev$, we have $f_V(e'v)>0$;
          \item for any proper subpath $ee'\subsetneq ev$, we have $f_V(ee')<0$;
          \item $e\in H$.
        \end{enumerate}
    \end{enumerate}
  \end{lemma}
  \begin{remark}
    Observe that the subset $V$ has no influence on the claim of (1) and the subset $H$ has no influence on the claim of (2).
  \end{remark}
  \begin{proof}
    We prove (1), the proof of (2) is similar.
    To see (a), note that as $e'$ traverses the path $he$ from $h$ to $e$, the value of $f_H(he')$ either increases by $c$, stays constant, or decreases by 1.
    Since $h\in H$, we have $f_H(hh)=c>0$ and so the value of $f_H(he')$ stays positive until it first becomes 0 when $e'=e$.
    Part (b) then follows from the additivity of $f_H$ and part (c) is the case when $e'=e$ here.
  \end{proof}

  \begin{corollary}
    Consider $H\sqcup V\subset D$ with $H\subset D_1$ and $V\subset D_2$.
    \begin{enumerate}
      \item For $h\in H$ and a vertical edge $v\in D(h;H)$, the condition (HGC) is not satisfied for the pair $(h,v)$ but for any $V\in\cC_D(H)$ the condition (VGC) is satisfied for the pair $(h,v)$.
        In particular, $D(v;V)$ is a proper subpath of $hv$ for any $V\in\cC_D(H)$.
      \item For $v\in V$ and a horizontal edge $h\in D(v;V)$, the condition (VGC) is not satisfied for the pair $(h,v)$ but for any $H\in\cC_D(V)$ the condition (HGC) is satisfied for the pair $(h,v)$.
        In particular, $D(h;H)$ is a proper subpath of $hv$ for any $H\in\cC_D(V)$.
    \end{enumerate}
  \end{corollary}
  \begin{proof}
    We prove (1), the proof of (2) is similar.
    By Lemma~\ref{le:shadow statistics}, $f_H(he)>0$ for every proper subpath of $D(h;H)=hv$ and so the condition (HGC) is not satisfied for the path $hv$, but then, since $V\in\cC_D(H)$, the condition (VGC) must be satisfied.
  \end{proof}

@@

    For $a_1,a_2\in\ZZ_{\ge0}$ with $(-a_2,-a_1+ca_2)\in\cI'$, define the \emph{dual dominant Dyck path} $\vv{D}^{dom}=\vv{D}^{dom}[a_1,a_2]$ to be the maximal Dyck path starting from $(0,0)$ and terminating at $(a_1,a_2)$ while staying below the lines $y=\frac{bc+\sqrt{bc(bc-4)}}{2c}x$ and $y-a_2=\frac{bc-\sqrt{bc(bc-4)}}{2c}(x-a_1)$.
  \end{definition}
  \begin{remark}
    \label{rem:intersection}
    The lines defining $D^{dom}[a_1,a_2]$ meet at the point 
    \[m_{a_1,a_2}:=\left( \frac{2ba_2-\big(bc-\sqrt{bc(bc-4)}\big)a_1}{2\sqrt{bc(bc-4)}} , \frac{\big(bc+\sqrt{bc(bc-4)}\big)a_2-2ca_1}{2\sqrt{bc(bc-4)}} \right).\]
  \end{remark}

@@

  \begin{theorem}
    \label{th:dominant Dyck path recursion}
    For $a_1,a_2\in\ZZ_{\ge0}$ with $(-a_1+ba_2,-a_2)\in\cI$, we have $\mu_{V,b}\big(D^{dom}[a_1,a_2]\big)=\vv{D}^{dom}[a_2,ba_2-a_1]$ and $\mu_{H,c}\big(D^{dom}[a_1,a_2]\big)=\vv{D}^{dom}[ca_1-a_2,a_1]$.
  \end{theorem}
  \begin{proof}
    We prove the result for $\mu_{H,c}$ as the result for $\mu_{V,b}$ immediately follows from $\mu_{H,b}$ being its inverse.
    Write $D'$ for the Dyck path $\mu_{H,c}\big(D^{dom}[a_1,a_2]\big)$ and observe that by construction $D'$ has $ca_1-a_2$ horizontal edges and $a_1$ vertical edges.
    Write $v'_1,\ldots,v'_{a_1}$ for the vertical edges of $D'$ with $v'_r$ immediately preceded by $\ell_r$ horizontal edges.
    We have to show (i) that every lattice point on $D'$ stays below the lines 
    \[y=\frac{bc+\sqrt{bc(bc-4)}}{2c}x \qquad \text{and} \qquad y-a_1=\frac{bc-\sqrt{bc(bc-4)}}{2c}\big(x-(ca_1-a_2)\big)\]
    while (ii) any lattice point above $D'$ is also above one of these lines.
